---
title: "Homework 04: Data frames and data wrangling"
format: html
editor: visual
---

Please read the entire chapter on data transformation from\
[R for Data Science](https://r4ds.hadley.nz/data-transform) before starting this homework.

This homework relies on the `nycflights13` package, which contains several data frames, including:

-   `airlines`
-   `airports`
-   `flights`
-   `planes`
-   `weather`

Loading `nycflights13` puts these data frames on the search path.

------------------------------------------------------------------------

## Setup

### Load packages (do this every time)

```{r}
library(tidyverse)
library(nycflights13)
```

Note - if you don't have these installed you will need to first install.packages('nycflights13')

------------------------------------------------------------------------

# Question 1: Filtering

Make a plot of **air time vs distance** (air time on the y-axis, distance on the x-axis) for all flights that meet the following criteria:

-   originate from LaGuardia airport (`"LGA"`)
-   departed on the 16th of the month
-   have a flight distance of less than 2000 miles

## Your code

```{r}
filtered_flights <- flights %>%
  filter(
    origin == "LGA",
    day == 16,
    distance < 2000
  ) %>%
  drop_na(air_time, distance)

ggplot(filtered_flights, aes(x = distance, y = air_time)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Air Time vs Distance",
    subtitle = "The Graph",
    x = "Distance (miles)",
    y = "Air Time (minutes)"
  ) +
  theme_minimal()
```

## Brief written response

In 2–4 sentences, describe what you observe in the plot. (For example: Is the relationship roughly linear? Are there any clear outliers?)

**Your answer here:**The graph shows a clear positive relationship between distance and air time. As distance increases, air time also increases in a mostly straight-line pattern.There are only a few small outlines, but overall the pattern is very consistent.

------------------------------------------------------------------------

# Question 2: Dealing with NAs

Make a data frame of all rows of `flights` that have values for *both* `arr_time` and `dep_time` (i.e., neither value is `NA`).

## Your code

```{r}
library(tidyverse)
library(nycflights13)

flights_complete <- flights %>%
  filter(!is.na(arr_time), !is.na(dep_time))

head(flights_complete)

```

## Filtering NAs (conceptual)

`ggplot()` will automatically remove `NA` values from a plot, but it emits a warning message about it. You *could* silence warnings using chunk options, but instead:

### Brief written response

Explain (in words) how you could prevent those `NA` values from appearing in the plot in the first place.

**Your answer here:**To prevent NA values from appearing in the plot, I would filter them out before making the graph. I can use ***`filter()`*** together with ***`!is.na()`*** to remove rows where important variables have missing values. This way, ggplot only receives clean data and won’t show a warning message.

------------------------------------------------------------------------

# Question 3: Adding columns

Create a data frame of **average flight speeds**, based on `air_time` and `distance`.

Then make either:

-   a histogram, **or**
-   a density plot

If you like, you may break the data out (e.g., by airline or another variable) in a way that you think makes sense.

## Your code

```{r}
library(tidyverse)
library(nycflights13)

speed_df <- flights %>%
  filter(!is.na(air_time), !is.na(distance)) %>%
  mutate(speed = distance / (air_time / 60))

ggplot(speed_df, aes(x = speed)) +
  geom_density() +
  labs(
    title = "Density Plot",
    x = "Speed (miles per hour)",
    y = "Density"
  ) +
  theme_minimal()


```

## Brief written response

Describe the main features of your speed distribution.

**Your answer here:**The density plot shows that most flights have an average speed around 400–450 miles per hour. The curve is slightly skewed to the right, which means a few flights are much faster than most.the speeds are fairly consistent for the majority of flights, with only a small number of unusually slow or fast flights

------------------------------------------------------------------------

# Rendering and submission (GitHub)

Canvas contains the course GitHub link and instructions for **forking** the course repository.

In this file, your job is to:

1.  Complete the homework by adding code + written responses above.
2.  Render this Quarto file to **HTML or PDF**.
3.  Move the rendered file(s) to the /doc folder (note that, if you render to html you will also have to move your hmk_04_data_frames_files folder)
4.  Add/commit both:
    -   this edited `.qmd` file
    -   the rendered output file(s) (HTML or PDF)
5.  Push your changes to your fork on GitHub.
6.  Open a Pull Request back to the course repository.

## What to submit on Canvas

On Canvas, submit a link to your GitHub fork repository (not files).
