---
title: "Homework 04: Data frames and data wrangling"
format: html
editor: visual
---

Please read the entire chapter on data transformation from\
[R for Data Science](https://r4ds.hadley.nz/data-transform) before starting this homework.

This homework relies on the `nycflights13` package, which contains several data frames, including:

-   `airlines`
-   `airports`
-   `flights`
-   `planes`
-   `weather`

Loading `nycflights13` puts these data frames on the search path.

------------------------------------------------------------------------

## Setup

### Load packages (do this every time)

```{r}
library(tidyverse)
library(nycflights13)
```

Note - if you don't have these installed you will need to first install.packages('nycflights13')

------------------------------------------------------------------------

# Question 1: Filtering

Make a plot of **air time vs distance** (air time on the y-axis, distance on the x-axis) for all flights that meet the following criteria:

-   originate from LaGuardia airport (`"LGA"`)
-   departed on the 16th of the month
-   have a flight distance of less than 2000 miles

## Your code

```{r}
# Filter flights according to the criteria above,
# then make the required plot.

flights_q1 <- flights %>%
  filter(
    origin == "LGA",
    day == 16,
    distance < 2000
  )

ggplot(flights_q1, aes(x = distance, y = air_time)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Air time vs distance (LGA flights on day 16; distance < 2000 miles)",
    x = "Distance (miles)",
    y = "Air time (minutes)"
  )

```

## Brief written response

In 2–4 sentences, describe what you observe in the plot. (For example: Is the relationship roughly linear? Are there any clear outliers?)

**Your answer here:**

The plot shows a strong positive roughly linear relationship between distance and air time. As flight distance increases, air time increases consistently. The points fall into visible vertical clusters at certain distances, which can correspond to common routes between specific airports. There is a little bit of spread within each cluster and a few flights with longer airtimes than expected but no extreme outliers.

------------------------------------------------------------------------

# Question 2: Dealing with NAs

Make a data frame of all rows of `flights` that have values for *both* `arr_time` and `dep_time` (i.e., neither value is `NA`).

## Your code

```{r}
# Create a new data frame that removes rows where arr_time or dep_time is NA.

flights_no_na_times <- flights %>%
  filter(!is.na(arr_time), !is.na(dep_time))

flights_no_na_times
```

## Filtering NAs (conceptual)

`ggplot()` will automatically remove `NA` values from a plot, but it emits a warning message about it. You *could* silence warnings using chunk options, but instead:

### Brief written response

Explain (in words) how you could prevent those `NA` values from appearing in the plot in the first place.

**Your answer here:**

To prevent NA values from appearing in the plot, we can filter them out of the dataset before plotting (ex: filter(!is.a(variable)) for any variables used in the plot). More generally, we can make a clean plotting data set that excludes rows where any mapped x/y (or grouping) columns are NA, so ggplot never receives missing values to drop.

------------------------------------------------------------------------

# Question 3: Adding columns

Create a data frame of **average flight speeds**, based on `air_time` and `distance`.

Then make either:

-   a histogram, **or**
-   a density plot

If you like, you may break the data out (e.g., by airline or another variable) in a way that you think makes sense.

## Your code

```{r}
# Create a new column for average speed.
# (Hint: think carefully about units — air_time is in minutes.)
# Then make a histogram or density plot.

speed_df <- flights %>%
  filter(!is.na(air_time), !is.na(distance), air_time > 0) %>%
  mutate(avg_speed_mph = distance * 60 / air_time)

ggplot(speed_df, aes(x = avg_speed_mph)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of average flight speeds",
    x = "Average speed (mph)",
    y = "Count of flights"
  )
```

## Brief written response

Describe the main features of your speed distribution.

**Your answer here:**

Most flights fall in a similar speed range, with the highest number around roughly 350-450 mph. There are fewer flights at very slow or very fast speeds, so the graph tapers off on both sides. Overall, the speeds look pretty consistent, with only a small number of flights outside the main range.

------------------------------------------------------------------------

# Rendering and submission (GitHub)

Canvas contains the course GitHub link and instructions for **forking** the course repository.

In this file, your job is to:

1.  Complete the homework by adding code + written responses above.
2.  Render this Quarto file to **HTML or PDF**.
3.  Move the rendered file(s) to the /doc folder (note that, if you render to html you will also have to move your hmk_04_data_frames_files folder)
4.  Add/commit both:
    -   this edited `.qmd` file
    -   the rendered output file(s) (HTML or PDF)
5.  Push your changes to your fork on GitHub.
6.  Open a Pull Request back to the course repository.

## What to submit on Canvas

On Canvas, submit a link to your GitHub fork repository (not files).
